---
title: "Loop graphs in RMarkdown using medicaldata"
author: "Zoë Turner"
date: "04/09/2021"
output: html_document
---

# Original RMarkdown

This example is a copy of the original by [Chris Beeley](https://github.com/nhs-r-community/demos-and-how-tos/blob/main/rmarkdown/loop_graphs/loop_dataframe_graphs.Rmd) which uses the {palmerpenguins} package. Chris and I were contacted by [Peter Higgins](https://twitter.com/ibddoctor) after our discussions about this on [Twitter](https://twitter.com/Letxuga007/status/1433729592896991261) and he shared the code for the package {medicaldata} using the same format Chris had shared. This is is his code within the original RMarkdown layout that Chris had created. 

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# install.package("medicaldata") # uncomment to load package from CRAN

library(medicaldata)
library(tidyverse)

# load the covid_testing data from the package

covid <- medicaldata::covid_testing

# make several datasets

covid_ed <- covid %>% 
  filter(clinic_name == "emergency dept")

covid_inpt_a <- covid %>% 
  filter(clinic_name == "inpatient ward a")

covid_urgent_care <- covid %>% 
  filter(clinic_name == "urgent care center")

lots_of_tests <- list(covid_ed, covid_inpt_a, covid_urgent_care)

```

Sometimes you may wish to run the same plotting function on multiple datasets. There's a simple way and a slightly-weird-but-works way.

Note that this is a contrived example and that this is not the simplest way of doing this particular thing.

### Simple

For this data (unlike for {palmerpenguins}) there are warnings about rows being deleted because of missing values.

```{r}

walk(lots_of_tests, function(x) {
  
  p <- x %>% 
    ggplot(aes(ct_result, pan_day)) +
    geom_point() +
    labs(x = "PCR Result\n(lower threshold cycle # means higher viral load",
         y = "Pandemic Day")
  
  print(p)
})
```

### Add a title

```{r}

testgroup_names <- c("Covid tests in ED", "Covid tests in inpatient Ward A", "Covid tests in Urgent Care")

walk2(lots_of_tests, testgroup_names, function(x, y) {
  
  p <- x %>% 
    ggplot(aes(ct_result, pan_day)) +
    geom_point() + 
    ggtitle(y) +
    labs(x = "PCR Result\n(lower threshold cycle # means higher viral load)",
         y = "Pandemic Day")
  
  print(p)
})

```

