---
title: "Demographic report for `r params$country`"
author: "Jon Minton"
date: "15/03/2022"
output: pdf
params: 
  code: "GBR_SCO"
  country: "Scotland"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse)


this_lt <- all_lt %>% 
  filter(code == params$code)

other_lt <- all_lt %>% 
  filter(!(code == params$code))

min_year <- min(this_lt$Year)
max_year <- max(this_lt$Year)

```

# Introduction

This is an attempt at a template for generating multiple markdown reports using parameters. 

It is based on the information on [the relevant section](https://bookdown.org/yihui/rmarkdown-cookbook/parameterized-reports.html) of the [rmarkdown cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/) website. 


# Introduction

`r params$country`, represented in the HMD by the code `r params$code`, has available data for the years `r min_year` to `r max_year`. Its life expectancy at birth over time is shown in red below, with other countries shown in a muted grey for comparison. 


```{r, echo = FALSE, fig.width = 8, fig.height = 6}
this_lt %>% 
  filter(Age == 0) %>% 
  ggplot(aes(x = Year, y = ex)) + 
  geom_line(
    aes(x = Year, y = ex, group = code), 
    color = "lightgray", inherit.aes = FALSE, 
    data = other_lt %>% 
      filter(Age == 0)
  ) + 
  geom_line(color = "darkred", size = 1.2) + 
  facet_wrap(~sex) +
  labs(x = "Year", y = "Life expectancy at birth",
       title = glue::glue("Life expectancy at birth for {params$country}")
  )


```



